---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# simtte

<!-- badges: start -->
<!-- badges: end -->

The goal of simtte is to fast simulate simple and complex time-to-event models also known as survival analysis. It also allows user-defined time-to-event models via the awesome package [**mrgsolve**](https://github.com/metrumresearchgroup/mrgsolve).

## Installation

You can install the released version of simtte from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("simtte")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("csetraynor/simtte")
```
## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(simtte)
# Load M-splines data example
data("ms_data")
mu <- ms_data$mu
basis <- ms_data$basis
coefs <- ms_data$coefs
time <- ms_data$time
# Simulate prognostic index, linear predictor
lp <- matrix(runif(nrow(basis)),  nrow = nrow(basis))

# simulate M-splines model
sim_tte(pi = lp, mu = mu, basis = basis, coefs = coefs, time = time, type = "ms", end_time = 100)
# simulate Weibull model
sim_tte(pi = lp, mu = -1, coefs = 1.1, time = time, type = "weibull", obs.only = F, obs.aug = T, delta = 0.05, end_time = 100)
```


