---
title: "Advanced Simulation Scenarios"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Simulation Scenarios}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(simtte)
```

## Exploring the Prognostic Index

The `explore_pi_tq_surv()` function allows you to understand how the
prognostic index (linear predictor) affects survival at a given quantile.

```{r explore-pi, eval = FALSE}
data_sim <- explore_pi_tq_surv(
  pi = seq(-3, 3, by = 0.1),
  mu = -1,
  shape = seq(0.9, 1.1, by = 0.1),
  end_time = 200,
  type = "weibull"
)
head(data_sim)
```

### Visualisation

```{r plot-pi, eval = FALSE}
library(ggplot2)
ggplot(data_sim, aes(x = exp(lp), y = survdiff_tq)) +
  geom_line(aes(color = factor(shape), group = shape)) +
  scale_x_log10() +
  labs(
    x = "Hazard Ratio",
    y = expression(Delta ~ "Survival at" ~ t[50]),
    color = "Shape"
  ) +
  geom_vline(xintercept = 1, linetype = 2) +
  geom_hline(yintercept = 0, linetype = 2) +
  theme_bw()
```

## Custom mrgsolve Models with sim_tte_df

If you have a custom mrgsolve model that outputs a survival probability
column, you can use `sim_tte_df()` directly to perform inverse transform
sampling on the output:
 
```{r custom, eval = FALSE}
# Assume 'mrg_output' is the output of an mrgsim() call with a 'p11' column
# result <- sim_tte_df(mrg_output, surv_var = "p11", id_var = "ID")
```

## Multiple Treatment Arms

You can simulate different treatment arms by specifying different
prognostic indices:

```{r arms, eval = FALSE}
set.seed(42)
n_per_arm <- 50
times <- seq(0.1, 50, by = 0.1)

# Control arm
lp_ctrl <- matrix(rep(0, n_per_arm), nrow = n_per_arm)
ctrl <- sim_tte(pi = lp_ctrl, mu = -1, coefs = 1.1,
  time = times, type = "weibull", end_time = 50)
ctrl$arm <- "Control"

# Treatment arm (lower hazard)
lp_trt <- matrix(rep(-0.5, n_per_arm), nrow = n_per_arm)
trt <- sim_tte(pi = lp_trt, mu = -1, coefs = 1.1,
  time = times, type = "weibull", end_time = 50)
trt$arm <- "Treatment"

combined <- rbind(ctrl, trt)
head(combined)
```
