---
title: "Bring Your Own PK/PD Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bring Your Own PK/PD Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(simtte)
```

`sim_tte_ode()`'s six built-in PK/PD models
(`vignette("pkpd-time-to-event")`) cover common structures; `tte_model()`
converts *any* `mrgsolve` model with an `$ODE`/`$DES` block into one
usable the same way. This vignette covers the conversion mechanics, how
it interacts with a model's own between-subject variability, the
boundary between a covariate effect and a model parameter, and the
compile-cost/caching trade-off.

## 1. A minimal conversion

Here is a complete, short model -- one-compartment PK with first-order
absorption -- with no survival scaffolding of its own:

```{r tte-model-source}
my_model <- "
$PARAM CL = 1, V = 20, KA = 1
$CMT DEPOT CENT
$ODE
dxdt_DEPOT = -KA*DEPOT;
dxdt_CENT = KA*DEPOT - (CL/V)*CENT;
"
```

Convert it, naming the hazard as a plain C++ expression in terms of
the model's own quantities (`CENT`, `V`) and new hazard parameters
(`H0`, `beta`):

```{r tte-model-convert}
tm <- tte_model(my_model, hazard = "H0 * exp(lp + beta * (CENT/V))",
  params = list(H0 = 0.01, beta = 0.05))
tm
```

`tm$code` is the full converted model, ready to inspect:

```{r tte-model-code}
cat(tm$code, sep = "\n")
```

Pass `tm` to `sim_tte_ode()` exactly like a built-in model name:

```{r tte-model-sim}
dose <- data.frame(ID = 1:30, time = 0, cmt = 1, amt = 100, evid = 1)
sim_tm <- sim_tte_ode(model = tm, n = 30, end = 20, delta = 2, data = dose,
  seed = 1)
head(sim_tm$events)
```

A bare compiled `mrgsolve` model (not wrapped by `tte_model()`) is not
accepted directly by `sim_tte_ode()` -- convert it first. This keeps
there to exactly one way to bring a model in.

## 2. What conversion actually does

`tte_model()` writes the same survival scaffold every built-in
`*_hazard` model already has -- a `p11` compartment with
`dxdt_p11 = -p11 * HAZ`, a per-subject uniform draw `U`, an
administrative horizon `END`, and the in-solver latch that records the
first crossing time -- onto your model, mechanically:

1. Your model's own `$ODE`/`$DES` content is required (a
   `$PKMODEL`/`$PRED`-only model is rejected, since the survival
   compartment must be co-integrated in the same ODE system; use an
   ODE-form `mrgsolve::modlib()` equivalent instead, e.g. `"pk2cmt"`
   not `"pk2"`).
2. `lp`, `U`, and `END` are added as new `$PARAM` entries (defaults 0,
   0, 1e9) -- always, whether or not your `hazard` expression uses
   `lp`. Include `lp` in `hazard` if `sim_tte_ode()`'s
   `covariates`/`beta`/`formula` mechanism should affect this model.
3. Your `hazard` expression becomes `HAZ`, computed inside a brand-new
   `$ODE` block appended after your model's own -- your model's own
   `$ODE`/`$DES` content is never edited.
4. `p11 = 1` is added as a new `$INIT` entry, and `p11` as a new
   compartment.
5. The event-detection latch (`TEVT`, `event_found`, and the bracket
   columns `T_PRE`/`P_PRE`/`P_POST` that `sim_tte_ode()`'s refinement
   step reads) is added to `$GLOBAL`/`$MAIN` -- the only two blocks
   edited in place, since `mrgsolve` allows only one of each.
6. New `$CAPTURE` entries expose the scaffold's own columns.
7. A reserved-name collision check runs before compiling: your model
   must not already declare `p11`, `U`, `END`, `HAZ`, `TEVT`,
   `event_found`, `T_PRE`, `P_PRE`, `P_POST`, or `lp` for anything
   else.
8. The result compiles via `mrgsolve::mcode_cache()` into a
   `"simtte_model"` object, ready for `sim_tte_ode(model = ...)`.

Verified to reproduce all six built-in hazard models exactly when
converting their own `mrgsolve::modlib()` backbones directly.

## 3. `$OMEGA` coexistence

If your own model already declares its own `$OMEGA` block (the classic
`TVCL`/`ETA()` idiom), `tte_model()` leaves it untouched and
`sim_tte_ode()`'s `omega` argument goes through it directly instead of
the `idata` route `vignette("pkpd-time-to-event")` shows for the
built-in models:

```{r tte-model-omega}
my_model_omega <- "
$PARAM TVCL = 1, TVV = 20, KA = 1
$OMEGA @labels ETA_CL ETA_V
0 0
$CMT DEPOT CENT
$MAIN
double CL = TVCL*exp(ETA_CL);
double V  = TVV*exp(ETA_V);
$ODE
dxdt_DEPOT = -KA*DEPOT;
dxdt_CENT = KA*DEPOT - (CL/V)*CENT;
"
tm_omega <- tte_model(my_model_omega,
  hazard = "H0 * exp(lp + beta * (CENT/V))",
  params = list(H0 = 0.01, beta = 0.05))
sim_tm_omega <- sim_tte_ode(model = tm_omega, omega = diag(c(0.09, 0.16)),
  n = 30, data = dose, end = 20, delta = 2, seed = 1)
head(sim_tm_omega$events)
```

`tte_model()`'s own `bsv_targets` argument is the alternative, for a
model with *no* declared `$OMEGA` block: it names which of the model's
structural parameters `sim_tte_ode()`'s `omega` may target via the
`idata` route instead, exactly like the built-in PK/PD models.
Supplying `bsv_targets` for a model that already declares `$OMEGA` is
ignored (with a `message()`) -- the two routes never combine for the
same model.

## 4. The `lp`/hazard boundary

A covariate's effect (`sim_tte_ode()`'s `covariates`/`beta`/`formula`)
must be linear in `beta`; a link that is nonlinear in a *parameter* you
want to estimate or vary belongs directly in `hazard` instead, composed
with whatever PK/PD state, BSV draw, or dosing your model already
provides:

```{r hazard-boundary}
tm_emax <- tte_model(my_model, hazard = "H0 * (CENT / (EC50 + CENT))",
  params = list(H0 = 0.01, EC50 = 5))
sim_emax <- sim_tte_ode(model = tm_emax, n = 30, end = 20, delta = 2,
  data = dose, seed = 1)
head(sim_emax$events)
```

Here `EC50` is a `param` you can override or, in a future variation,
sample per subject via `idata` -- something a covariate column can
never do, since a covariate's effect only ever multiplies a fixed
column by a coefficient. See `vignette("pkpd-time-to-event")` section 3
for the same boundary from the covariate side, including a
demonstration that a *fixed* `EC50` gives identical results whichever
way it is written.

## 5. Compile cost and caching

Every `tte_model()` conversion is a real C++ compile (a few seconds).
`mrgsolve::mcode_cache()` (used internally) skips recompiling an
unchanged, previously-seen model within the same cache directory --
`options(mrgsolve.project = <a persistent directory>)` makes that
cache survive across R sessions too, worthwhile if you reconvert the
same model many times (e.g. a simulation study script re-run
repeatedly). Two caveats: a caller-supplied `name` reused for
genuinely different `hazard`/`params` content, called back to back,
can in principle collide within the same build directory (leave
`name` at its default, which avoids this by construction, since it is
derived from a hash of the converted code); and in a
parallel workflow, give each worker its own cache directory, or simply
let each one recompile -- the compile cost is small next to the
parallel speed-up.
