---
title: "Joint PK/PD and Time-to-Event Simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Joint PK/PD and Time-to-Event Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(simtte)
```

## 1. What this is for

**simtte** simulates time-to-event data where the hazard is driven by
a pharmacokinetic/pharmacodynamic (PK/PD) model rather than a fixed
parametric shape. The survival probability is added to the PK/PD
model as one more ODE compartment, `p11`, alongside the model's own
concentration/response compartments:

$$\frac{d\,p_{11}}{dt} = -p_{11}(t) \cdot \mathrm{HAZ}(t)$$

where `HAZ(t)` is any expression built from the model's own PK/PD
quantities (concentration, response, drug-target complex, ...). Each
subject gets one draw $U \sim \mathrm{Uniform}(0, 1)$; the ODE solver
is watched *during* integration for the first time at which
$p_{11}(t) \le U$, and that is the subject's event time (or the
subject is censored at the follow-up horizon `end` if no crossing
occurs). This happens inside the solver, not after the fact from a
pre-simulated trajectory -- the mechanism `sim_tte_ode()` implements.

Two ways to get a PK/PD model into this mechanism:

- Use one of six **built-in library models** (`sim_tte_ode(model = "pk_hazard")`,
  `"irm1_hazard"`-`"irm4_hazard"`, `"tmdd_hazard"`), each a well-known
  PK/PD structure with the hazard already wired in.
- Convert **your own** `mrgsolve` PK/PD model with `tte_model()`.

Everything below is real, executed code -- not illustrative snippets.

## 2. `sim_tte_ode()` with a baseline model

The simplest case has no PK/PD at all: a built-in Weibull hazard.

```{r weibull-baseline}
sim <- sim_tte_ode(model = "weibull", param = list(mu = -1, shape = 1.5),
  n = 30, end = 20, delta = 2, seed = 1)
head(sim$events)
```

`sim$events` is always three columns: `ID`, `sim_time`, `sim_status`
(1 = event, 0 = censored at `end`).

A covariate that changes over time (`covariates`/`beta`) composes with
any model, including this one:

```{r weibull-covariate}
cov <- data.frame(time = c(0, 10), sex = c(0, 1))
sim_cov <- sim_tte_ode(model = "weibull", param = list(mu = -1, shape = 1.5),
  n = 30, end = 20, delta = 2, covariates = cov, beta = c(sex = 0.8),
  seed = 1)
head(sim_cov$events)
```

`lp(t) = beta["sex"] * sex(t)` is folded into the hazard from `time =
10` onward for every subject; every subject's `covariates` must
include an observation at `time = 0` (no implicit `lp(0)`).

## 3. A PK/PD library model, with dosing and between-subject variability

`irm1_hazard` links the hazard to an indirect-response PD compartment
relative to its own pre-dose baseline. Dosing is ordinary `mrgsolve`
event data, passed via `data`:

```{r irm1-dose}
dose <- data.frame(ID = 1:30, time = 0, cmt = 1, amt = 100, evid = 1)
sim_pkpd <- sim_tte_ode(model = "irm1_hazard",
  param = list(H0 = 0.01, beta_r = 1), n = 30, end = 20, delta = 2,
  data = dose, seed = 1)
head(sim_pkpd$events)
```

Between-subject variability (BSV) on a structural PK/PD parameter
(here, clearance `CL`) is `omega`, a named matrix (any subset of the
model's own targets -- see `?sim_tte_ode` "Between-subject
variability" for the full list per model):

```{r irm1-omega}
om <- matrix(0.09, dimnames = list("CL", "CL"))
sim_bsv <- sim_tte_ode(model = "irm1_hazard", omega = om,
  param = list(H0 = 0.01, beta_r = 1), n = 30, end = 20, delta = 2,
  data = dose, seed = 1)
head(sim_bsv$events)
```

`sim_bsv$events` does not echo back each subject's own drawn `CL` --
it is consumed internally by the ODE solver, not one of the reported
columns. If you want to see the per-subject values, build `idata`
yourself instead of using `omega` (this is exactly the mechanism
`omega` uses internally, exposed as an ordinary R data frame):

```{r irm1-idata}
set.seed(1)
n <- 30
CL_i <- 1 * exp(rnorm(n, 0, sqrt(0.09)))  # population CL = 1, log-normal BSV
idata <- data.frame(ID = seq_len(n), CL = CL_i)
head(idata)

sim_idata <- sim_tte_ode(model = "irm1_hazard", idata = idata,
  param = list(H0 = 0.01, beta_r = 1), end = 20, delta = 2, data = dose,
  seed = 1)
head(sim_idata$events)
```

Any `idata` column matching a declared model parameter overrides that
parameter per subject -- `U`/`END` (drawn/filled automatically when
absent) work the same way.

## 4. `tte_model()`: bring in your own PK/PD model

The six library models cover common structures; `tte_model()` converts
*any* `mrgsolve` model with an `$ODE`/`$DES` block. Here is a complete,
short model -- one-compartment PK with first-order absorption -- with
no survival scaffolding of its own:

```{r tte-model-source}
my_model <- "
$PARAM CL = 1, V = 20, KA = 1
$CMT DEPOT CENT
$ODE
dxdt_DEPOT = -KA*DEPOT;
dxdt_CENT = KA*DEPOT - (CL/V)*CENT;
"
```

Convert it, naming the hazard as a plain C++ expression in terms of
the model's own quantities (`CENT`, `V`) and new hazard parameters
(`H0`, `beta`):

```{r tte-model-convert}
tm <- tte_model(my_model, hazard = "H0 * exp(lp + beta * (CENT/V))",
  params = list(H0 = 0.01, beta = 0.05))
tm
```

`tm$code` is the full converted model, ready to inspect:

```{r tte-model-code}
cat(tm$code, sep = "\n")
```

Pass `tm` to `sim_tte_ode()` exactly like a built-in model name:

```{r tte-model-sim}
sim_tm <- sim_tte_ode(model = tm, n = 30, end = 20, delta = 2, data = dose,
  seed = 1)
head(sim_tm$events)
```

If your own model already declares its own `$OMEGA` block (the classic
`TVCL`/`ETA()` idiom), `tte_model()` leaves it untouched and
`sim_tte_ode()`'s `omega` argument goes through it directly instead of
the `idata` route shown in section 3:

```{r tte-model-omega}
my_model_omega <- "
$PARAM TVCL = 1, TVV = 20, KA = 1
$OMEGA @labels ETA_CL ETA_V
0 0
$CMT DEPOT CENT
$MAIN
double CL = TVCL*exp(ETA_CL);
double V  = TVV*exp(ETA_V);
$ODE
dxdt_DEPOT = -KA*DEPOT;
dxdt_CENT = KA*DEPOT - (CL/V)*CENT;
"
tm_omega <- tte_model(my_model_omega,
  hazard = "H0 * exp(lp + beta * (CENT/V))",
  params = list(H0 = 0.01, beta = 0.05))
sim_tm_omega <- sim_tte_ode(model = tm_omega, omega = diag(c(0.09, 0.16)),
  n = 30, data = dose, end = 20, delta = 2, seed = 1)
head(sim_tm_omega$events)
```

A bare compiled `mrgsolve` model (not wrapped by `tte_model()`) is not
accepted directly by `sim_tte_ode()` -- convert it first. This keeps
there to exactly one way to bring a model in.

## 5. Accuracy, messages, and when things get slow

`delta` (and `add`) only control how finely the returned event time is
*reported* -- the event time itself is located inside the solver, at
its own internal step resolution, not snapped to the requested output
grid.

Two situations trigger a `message()` (never a change in what is
simulated, just a heads-up):

- **A very steep hazard at a coarse `delta`** can force the solver
  across a region where the hazard changes by orders of magnitude in
  one internal step; when that happens for some subjects, `sim_tte_ode()`
  falls back to a coarser-grid event-time estimate for just those
  subjects and says so, naming which of two known causes applies and
  which of `delta` or `rtol`/`atol` fixes it.
- **Weibull `shape < 0.05`**: the ODE form of the Weibull hazard is
  mathematically singular as $t \to 0$ for small shapes; below 0.05,
  accuracy degrades near $t = 0$ specifically. The simulation still
  runs; `sim_tte(type = "weibull")` (the closed-form path) is exact
  there if that matters more than the in-solver mechanism's other
  properties.

```{r shape-guardrail}
msgs <- capture.output(
  sim_small_shape <- sim_tte_ode(model = "weibull",
    param = list(mu = -1, shape = 0.01), n = 5, end = 10, delta = 1,
    seed = 1),
  type = "message"
)
msgs
```

`model = "mspline"` evaluates a quadratic M-spline basis *inside* the
compiled model as a genuinely continuous hazard, at the cost of a
fixed set of shipped interior-knot counts (3, 5, or 7); `sim_tte(type
= "ms")` accepts an arbitrary knot count but treats the hazard as
piecewise-constant on its own reporting grid instead. Use whichever
property you need more.

Every `tte_model()` conversion is a real C++ compile (a few seconds).
`mrgsolve::mcode_cache()` (used internally) skips recompiling an
unchanged, previously-seen model within the same cache directory --
`options(mrgsolve.project = <a persistent directory>)` makes that
cache survive across R sessions too, worthwhile if you reconvert the
same model many times (e.g. a simulation study script re-run
repeatedly). Two caveats: a caller-supplied `name` reused for
genuinely different `hazard`/`params` content, called back to back,
can in principle collide within the same build directory (leave
`name` at its default, which avoids this by construction); and in a
parallel workflow, give each worker its own cache directory, or simply
let each one recompile -- the compile cost is small next to the
parallel speed-up.

## 6. Right censoring

Every simulation so far has been administratively censored at `end`
only. `add_censoring()` layers a second, independent censoring process
on top of any `$events`-shaped output (`sim_tte_ode()`, `sim_tte()`, or
`sim_tte_df()`), drawing a per-subject censoring time and taking
`min(event time, censoring time, end)`:

```{r censoring-basic}
sim_uncensored <- sim_tte_ode(model = "exponential",
  param = list(H0 = 0.05), n = 200, end = 30, delta = 2, seed = 1)
table(sim_uncensored$events$sim_reason)

censored <- add_censoring(sim_uncensored$events,
  censoring = list(dist = "exponential", rate = 0.02), end = 30, seed = 2)
table(censored$sim_reason)
```

`sim_status`/`sim_time` are updated in place (so any existing code
reading those two columns keeps working); `sim_reason` records why --
`"event"`, `"censored"` (pulled earlier by the independent draw), or
`"administrative"` (hit `end`). `sim_tte_ode()` also accepts a
`censoring = ` argument directly, drawing inside the same seeded call
as `U`:

```{r censoring-arg}
sim_censored <- sim_tte_ode(model = "exponential", param = list(H0 = 0.05),
  n = 200, end = 30, delta = 2,
  censoring = list(dist = "exponential", rate = 0.02), seed = 1)
table(sim_censored$events$sim_reason)
```

`"exponential"`, `"weibull"` (fixed `shape`, solves for `scale`),
`"uniform"` (a fixed interval -- the common case for uniform-accrual
administrative censoring), `"lognormal"` (fixed `sdlog`), `"gamma"`
(fixed `shape`, `rate` only), and `"user"` (any function `n ->
numeric(n)`) are the six `dist` options; see `?add_censoring`.
Censoring that depends on a subject's own simulated PK/PD trajectory
(informative dropout, e.g. driven by toxicity) is out of scope -- see
`reports/16_censoring_design.md` for why, and what a future version
supporting it would need to change.

### Choosing the censoring rate for a target censoring fraction

**`censoring_rate_for()` solves for a `dist` parameter (rate, scale,
meanlog, or gamma rate) that would give a target independent-censoring
fraction, estimated from the event times of an already-run,
uncensored simulation.** It is an approximation, not an exact
calculation -- the estimate depends on the particular uncensored run
it is given, so always redraw on a larger cohort afterward and check
the realized fraction landed close to the target, the way this example
does:

```{r censoring-rate-for}
# 1. Solve: what rate gives ~20% independently censored, given this
#    model's own event-time scale?
event_times <- sim_uncensored$events$sim_time[
  sim_uncensored$events$sim_status == 1]
rate_20pct <- censoring_rate_for(event_times, target = 0.2)
rate_20pct

# 2. Redraw: apply that rate to a larger cohort.
check <- sim_tte_ode(model = "exponential", param = list(H0 = 0.05),
  n = 2000, end = 30, delta = 2,
  censoring = list(dist = "exponential", rate = rate_20pct), seed = 1)

# 3. Check: did it land close to 20%?
mean(check$events$sim_reason == "censored")
```

If the realized fraction in step 3 is not close enough, re-solve with a
larger or different uncensored sample rather than trusting the first
answer -- the check is the point, not an optional extra.

## 7. Interval censoring

Right censoring still gives an *exact* event time when an event
happens. In a real trial, many endpoints (progression, response) are
only checked at scheduled visits -- an event is known to have happened
somewhere between the last visit where the subject was event-free and
the first visit where it was detected. `add_interval_censoring()` maps
an already-simulated (and, if used, already right-censored) outcome
onto that interval:

```{r interval-basic}
visits <- c(0, 5, 10, 15, 20, 25, 30)
interval <- add_interval_censoring(sim_censored$events, visits = visits)
head(interval[, c("ID", "sim_time", "sim_status", "sim_reason",
  "sim_time_left", "sim_time_right")])
```

`sim_time`/`sim_status`/`sim_reason` are untouched -- `sim_time_left`/
`sim_time_right` are added on top, so anything already reading the
first three keeps working. `sim_time_right` is `Inf` whenever there is
no known upper bound: a censored subject (never detected as an event
under this schedule), or -- more subtly -- an event that genuinely
happened but only *after* the subject's last scheduled visit, so the
schedule itself never confirmed it. `sim_status` still reads `1` for
that second case; the exact columns are the simulation's ground truth,
independent of what the visit schedule happened to observe.

**Order matters**: apply `add_interval_censoring()` *after*
`add_censoring()`, never before -- a dropout changes `sim_time`/
`sim_status`, and the interval mapping needs to see the final value to
know there was no assessment after the dropout time. `sim_tte_ode()`'s
own `censoring =`/`visits =` arguments apply them in this order
automatically.

A visit schedule can also be per-subject (a `data.frame(ID, time)`), or
generated with fixed spacing plus random jitter -- realistic, since
scheduled visits rarely land exactly on the nominal day:

```{r interval-jitter}
sched <- visit_schedule(n = 200, every = 4, end = 24, jitter = 1, seed = 1)
head(sched)

sim_jittered <- sim_tte_ode(model = "exponential", param = list(H0 = 0.05),
  n = 200, end = 24, delta = 2, visits = list(every = 4, jitter = 1),
  seed = 1)
head(sim_jittered$events[, c("sim_time", "sim_status", "sim_time_left",
  "sim_time_right")])
```

For a nonparametric interval-censored survival fit, convert
`sim_time_right`'s `Inf` to `NA` (the `survival` package's own
convention for an open upper bound) and hand both bounds to
`survival::Surv(..., type = "interval2")`:

```{r interval-survfit}
right_for_surv <- ifelse(is.infinite(interval$sim_time_right), NA,
  interval$sim_time_right)
fit <- survival::survfit(survival::Surv(time = sim_time_left,
  time2 = right_for_surv, type = "interval2") ~ 1,
  data = data.frame(sim_time_left = interval$sim_time_left,
    right_for_surv = right_for_surv))
summary(fit, times = visits)
```

## 8. Backward compatibility

`sim_tte()` and `sim_tte_df()` -- the original, closed-form/grid-based
API -- are unchanged:

```{r legacy-api}
set.seed(42)
lp <- matrix(rnorm(20, 0, 0.5), nrow = 20)
result <- sim_tte(pi = lp, mu = -1, coefs = 1.1,
  time = seq(0.1, 20, by = 0.5), type = "weibull", end_time = 20)
head(result)
```

Reasons to still reach for them instead of `sim_tte_ode()`:

- **An arbitrary M-spline knot count** (`sim_tte(type = "ms")`) --
  `sim_tte_ode(model = "mspline")` ships only 3 fixed variants.
- **Exact Weibull behavior for very small shapes** (`shape < 0.05`,
  see section 5) -- the closed-form calculation has no singularity
  there.
- **An externally simulated `p11` trajectory** (from any source, not
  necessarily `mrgsolve`) -- `sim_tte_df()` accepts any validated
  `(ID, time, p11)` data frame directly; `sim_tte_ode()` always runs
  the ODE solver itself.
