---
title: "Joint PK/PD and Time-to-Event Simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Joint PK/PD and Time-to-Event Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(simtte)
```

## 1. What this is for

**simtte** simulates time-to-event data where the hazard is driven by
a pharmacokinetic/pharmacodynamic (PK/PD) model rather than a fixed
parametric shape. The survival probability is added to the PK/PD
model as one more ODE compartment, `p11`, alongside the model's own
concentration/response compartments:

$$\frac{d\,p_{11}}{dt} = -p_{11}(t) \cdot \mathrm{HAZ}(t)$$

where `HAZ(t)` is any expression built from the model's own PK/PD
quantities (concentration, response, drug-target complex, ...). Each
subject gets one draw $U \sim \mathrm{Uniform}(0, 1)$; the ODE solver
is watched *during* integration for the first time at which
$p_{11}(t) \le U$, and that is the subject's event time (or the
subject is censored at the follow-up horizon `end` if no crossing
occurs). This happens inside the solver, not after the fact from a
pre-simulated trajectory -- the mechanism `sim_tte_ode()` implements.

This vignette is the tour: the built-in models, covariates, and
between-subject variability, all through `sim_tte_ode()`. Two more
vignettes cover the rest of the toolkit in depth --
`vignette("bring-your-own-model")` for converting your own PK/PD model
with `tte_model()`, and `vignette("censoring-and-assessment")` for
right/interval censoring and realistic visit schedules. Everything
below is real, executed code, not illustrative snippets.

## 2. Library models

The simplest case has no PK/PD at all: a built-in Weibull hazard.

```{r weibull-baseline}
sim <- sim_tte_ode(model = "weibull", param = list(mu = -1, shape = 1.5),
  n = 30, end = 20, delta = 2, seed = 1)
head(sim$events)
```

`sim$events` is always at least three columns: `ID`, `sim_time`,
`sim_status` (1 = event, 0 = censored at `end`). `"exponential"`,
`"weibull"`, `"gompertz"`, and `"mspline"` (a flexible, continuously
evaluated baseline hazard) are the closed-hazard-shape models;
`sim_tte_ode_models()` lists every built-in name, including the six
PK/PD-linked ones below, together with what between-subject
variability each supports.

```{r sim-tte-ode-models}
sim_tte_ode_models()
```

Six more models link the hazard to a genuine PK/PD mechanism instead
of a fixed shape. `irm1_hazard` links it to an indirect-response PD
compartment relative to its own pre-dose baseline; dosing is ordinary
`mrgsolve` event data, passed via `data`:

```{r irm1-dose}
dose <- data.frame(ID = 1:30, time = 0, cmt = 1, amt = 100, evid = 1)
sim_pkpd <- sim_tte_ode(model = "irm1_hazard",
  param = list(H0 = 0.01, beta_r = 1), n = 30, end = 20, delta = 2,
  data = dose, seed = 1)
head(sim_pkpd$events)
```

The other five (`pk_hazard`, concentration-driven; `irm2_hazard`-`irm4_hazard`,
the other indirect-response types; `tmdd_hazard`, target-mediated drug
disposition) work the same way, each reusing an unedited `mrgsolve::modlib()`
PK/PD backbone plus a hazard link -- see `?sim_tte_ode` "PK/PD-linked
hazard models" for each one's exact hazard expression.

## 3. Covariates and the linear predictor

Covariates compose with any model via a formula and a design matrix:
`sim_tte_ode()` builds `X` with `stats::model.matrix()` and computes
`lp = X %*% beta`.

```{r weibull-covariate}
baseline <- data.frame(age = rnorm(30, 50, 10),
  arm = factor(rep(c("control", "treated"), 15)))
sim_cov <- sim_tte_ode(model = "weibull", param = list(mu = -1, shape = 1.5),
  n = 30, end = 20, delta = 2, covariates = baseline,
  formula = ~ age + arm, beta = c(age = 0.01, armtreated = -0.5),
  seed = 1)
head(sim_cov$events)
```

`baseline` has no `time` column, so it is a fixed, per-subject covariate
set (equivalent to a single observation at `time = 0`, in effect for
the whole follow-up -- a multi-row frame like this one also determines
`n` when it is not supplied); `arm` is coded as an ordinary treatment-contrast
factor, and the intercept `formula` would otherwise imply is dropped
automatically (with a message) since the model's own `mu` already plays
that role. `formula` accepts anything `stats::model.matrix()` does --
`I(age^2)`, `log()`, interactions, `splines::ns()` -- as long as the
result is linear in `beta`.

The original interface, one raw covariate column per named coefficient
with no formula at all, still works unchanged:

```{r weibull-covariate-legacy}
cov <- data.frame(time = c(0, 10), sex = c(0, 1))
sim_legacy <- sim_tte_ode(model = "weibull", param = list(mu = -1, shape = 1.5),
  n = 30, end = 20, delta = 2, covariates = cov, beta = c(sex = 0.8),
  seed = 1)
```

### Where a covariate ends and a model parameter begins

`formula`/`beta` only ever need to be linear in `beta`, not in the
covariates -- a covariate column can already be an arbitrarily
nonlinear, *fixed* transform (`I(age^2)`, or a column computed as
`dose / (EC50 + dose)` for a literal numeric `EC50`). A link that is
nonlinear in a *parameter* you want to estimate or vary -- an Emax
effect whose `EC50` is itself unknown -- does not fit here: it belongs
in the model's own hazard expression instead, via
`tte_model(hazard = "H0 * (CP / (EC50 + CP))", params = list(EC50 = 5))`
(see `vignette("bring-your-own-model")`), where `EC50` composes with
PK/PD states, BSV, and dosing directly, and can be varied like any
other model parameter. A *fixed* `EC50` gives the identical event-time
distribution written either way; only the second way lets `EC50` vary.

## 4. Between-subject variability

Between-subject variability (BSV) on a structural PK/PD parameter
(here, clearance `CL`) is `omega`, a named matrix (any subset of the
model's own targets -- see `?sim_tte_ode` "Between-subject
variability" for the full list per model):

```{r irm1-omega}
om <- matrix(0.09, dimnames = list("CL", "CL"))
sim_bsv <- sim_tte_ode(model = "irm1_hazard", omega = om,
  param = list(H0 = 0.01, beta_r = 1), n = 30, end = 20, delta = 2,
  data = dose, seed = 1)
head(sim_bsv$events)
```

`sim_bsv$events` does not echo back each subject's own drawn `CL` --
it is consumed internally by the ODE solver, not one of the reported
columns. If you want to see the per-subject values, build `idata`
yourself instead of using `omega` (this is exactly the mechanism
`omega` uses internally, exposed as an ordinary R data frame):

```{r irm1-idata}
set.seed(1)
n <- 30
CL_i <- 1 * exp(rnorm(n, 0, sqrt(0.09)))  # population CL = 1, log-normal BSV
idata <- data.frame(ID = seq_len(n), CL = CL_i)
head(idata)

sim_idata <- sim_tte_ode(model = "irm1_hazard", idata = idata,
  param = list(H0 = 0.01, beta_r = 1), end = 20, delta = 2, data = dose,
  seed = 1)
head(sim_idata$events)
```

Any `idata` column matching a declared model parameter overrides that
parameter per subject -- `U`/`END` (drawn/filled automatically when
absent) work the same way. A user-supplied model with its own declared
`$OMEGA` block uses `omega` the ordinary `mrgsolve` way instead -- see
`vignette("bring-your-own-model")`.

## 5. Beyond this tour

Two more vignettes cover the rest in depth:

- **`vignette("bring-your-own-model")`**: `tte_model()` converts *any*
  `mrgsolve` model with an `$ODE`/`$DES` block into one usable here,
  for when the six built-in PK/PD models above do not cover your
  mechanism.
- **`vignette("censoring-and-assessment")`**: right censoring
  (including choosing a rate for a target censoring fraction), interval
  censoring from a visit schedule (including left censoring), and
  realistic visit-schedule generators (jitter, missed visits/dropout,
  and outcome-reactive schedules).

## 6. Accuracy, messages, and when things get slow

`delta` (and `add`) only control how finely the returned event time is
*reported* -- the event time itself is located inside the solver, at
its own internal step resolution, not snapped to the requested output
grid.

Two situations trigger a `message()` (never a change in what is
simulated, just a heads-up):

- **A very steep hazard at a coarse `delta`** can force the solver
  across a region where the hazard changes by orders of magnitude in
  one internal step; when that happens for some subjects, `sim_tte_ode()`
  falls back to a coarser-grid event-time estimate for just those
  subjects and says so, naming which of two known causes applies and
  which of `delta` or `rtol`/`atol` fixes it.
- **Weibull `shape < 0.05`**: the ODE form of the Weibull hazard is
  mathematically singular as $t \to 0$ for small shapes; below 0.05,
  accuracy degrades near $t = 0$ specifically. The simulation still
  runs; `sim_tte(type = "weibull")` (the closed-form path) is exact
  there if that matters more than the in-solver mechanism's other
  properties.

```{r shape-guardrail}
msgs <- capture.output(
  sim_small_shape <- sim_tte_ode(model = "weibull",
    param = list(mu = -1, shape = 0.01), n = 5, end = 10, delta = 1,
    seed = 1),
  type = "message"
)
msgs
```

`model = "mspline"` evaluates a quadratic M-spline basis *inside* the
compiled model as a genuinely continuous hazard, at the cost of a
fixed set of shipped interior-knot counts (3, 5, or 7); `sim_tte(type
= "ms")` accepts an arbitrary knot count but treats the hazard as
piecewise-constant on its own reporting grid instead. Use whichever
property you need more.

## 7. Backward compatibility

`sim_tte()` and `sim_tte_df()` -- the original, closed-form/grid-based
API -- are unchanged:

```{r legacy-api}
set.seed(42)
lp <- matrix(rnorm(20, 0, 0.5), nrow = 20)
result <- sim_tte(pi = lp, mu = -1, coefs = 1.1,
  time = seq(0.1, 20, by = 0.5), type = "weibull", end_time = 20)
head(result)
```

Reasons to still reach for them instead of `sim_tte_ode()`:

- **An arbitrary M-spline knot count** (`sim_tte(type = "ms")`) --
  `sim_tte_ode(model = "mspline")` ships only 3 fixed variants.
- **Exact Weibull behavior for very small shapes** (`shape < 0.05`,
  see section 6) -- the closed-form calculation has no singularity
  there.
- **An externally simulated `p11` trajectory** (from any source, not
  necessarily `mrgsolve`) -- `sim_tte_df()` accepts any validated
  `(ID, time, p11)` data frame directly; `sim_tte_ode()` always runs
  the ODE solver itself.
