---
title: "Joint PK/PD and Time-to-Event Simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Joint PK/PD and Time-to-Event Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(simtte)
```

## 1. What this is for

**simtte** simulates time-to-event data where the hazard is driven by
a pharmacokinetic/pharmacodynamic (PK/PD) model rather than a fixed
parametric shape. The survival probability is added to the PK/PD
model as one more ODE compartment, `p11`, alongside the model's own
concentration/response compartments:

$$\frac{d\,p_{11}}{dt} = -p_{11}(t) \cdot \mathrm{HAZ}(t)$$

where `HAZ(t)` is any expression built from the model's own PK/PD
quantities (concentration, response, drug-target complex, ...). Each
subject gets one draw $U \sim \mathrm{Uniform}(0, 1)$; the ODE solver
is watched *during* integration for the first time at which
$p_{11}(t) \le U$, and that is the subject's event time (or the
subject is censored at the follow-up horizon `end` if no crossing
occurs). This happens inside the solver, not after the fact from a
pre-simulated trajectory -- the mechanism `sim_tte_ode()` implements.

Two ways to get a PK/PD model into this mechanism:

- Use one of six **built-in library models** (`sim_tte_ode(model = "pk_hazard")`,
  `"irm1_hazard"`-`"irm4_hazard"`, `"tmdd_hazard"`), each a well-known
  PK/PD structure with the hazard already wired in.
- Convert **your own** `mrgsolve` PK/PD model with `tte_model()`.

Everything below is real, executed code -- not illustrative snippets.

## 2. `sim_tte_ode()` with a baseline model

The simplest case has no PK/PD at all: a built-in Weibull hazard.

```{r weibull-baseline}
sim <- sim_tte_ode(model = "weibull", param = list(mu = -1, shape = 1.5),
  n = 30, end = 20, delta = 2, seed = 1)
head(sim$events)
```

`sim$events` is always three columns: `ID`, `sim_time`, `sim_status`
(1 = event, 0 = censored at `end`).

A covariate that changes over time (`covariates`/`beta`) composes with
any model, including this one:

```{r weibull-covariate}
cov <- data.frame(time = c(0, 10), sex = c(0, 1))
sim_cov <- sim_tte_ode(model = "weibull", param = list(mu = -1, shape = 1.5),
  n = 30, end = 20, delta = 2, covariates = cov, beta = c(sex = 0.8),
  seed = 1)
head(sim_cov$events)
```

`lp(t) = beta["sex"] * sex(t)` is folded into the hazard from `time =
10` onward for every subject; every subject's `covariates` must
include an observation at `time = 0` (no implicit `lp(0)`).

## 3. A PK/PD library model, with dosing and between-subject variability

`irm1_hazard` links the hazard to an indirect-response PD compartment
relative to its own pre-dose baseline. Dosing is ordinary `mrgsolve`
event data, passed via `data`:

```{r irm1-dose}
dose <- data.frame(ID = 1:30, time = 0, cmt = 1, amt = 100, evid = 1)
sim_pkpd <- sim_tte_ode(model = "irm1_hazard",
  param = list(H0 = 0.01, beta_r = 1), n = 30, end = 20, delta = 2,
  data = dose, seed = 1)
head(sim_pkpd$events)
```

Between-subject variability (BSV) on a structural PK/PD parameter
(here, clearance `CL`) is `omega`, a named matrix (any subset of the
model's own targets -- see `?sim_tte_ode` "Between-subject
variability" for the full list per model):

```{r irm1-omega}
om <- matrix(0.09, dimnames = list("CL", "CL"))
sim_bsv <- sim_tte_ode(model = "irm1_hazard", omega = om,
  param = list(H0 = 0.01, beta_r = 1), n = 30, end = 20, delta = 2,
  data = dose, seed = 1)
head(sim_bsv$events)
```

`sim_bsv$events` does not echo back each subject's own drawn `CL` --
it is consumed internally by the ODE solver, not one of the reported
columns. If you want to see the per-subject values, build `idata`
yourself instead of using `omega` (this is exactly the mechanism
`omega` uses internally, exposed as an ordinary R data frame):

```{r irm1-idata}
set.seed(1)
n <- 30
CL_i <- 1 * exp(rnorm(n, 0, sqrt(0.09)))  # population CL = 1, log-normal BSV
idata <- data.frame(ID = seq_len(n), CL = CL_i)
head(idata)

sim_idata <- sim_tte_ode(model = "irm1_hazard", idata = idata,
  param = list(H0 = 0.01, beta_r = 1), end = 20, delta = 2, data = dose,
  seed = 1)
head(sim_idata$events)
```

Any `idata` column matching a declared model parameter overrides that
parameter per subject -- `U`/`END` (drawn/filled automatically when
absent) work the same way.

## 4. `tte_model()`: bring in your own PK/PD model

The six library models cover common structures; `tte_model()` converts
*any* `mrgsolve` model with an `$ODE`/`$DES` block. Here is a complete,
short model -- one-compartment PK with first-order absorption -- with
no survival scaffolding of its own:

```{r tte-model-source}
my_model <- "
$PARAM CL = 1, V = 20, KA = 1
$CMT DEPOT CENT
$ODE
dxdt_DEPOT = -KA*DEPOT;
dxdt_CENT = KA*DEPOT - (CL/V)*CENT;
"
```

Convert it, naming the hazard as a plain C++ expression in terms of
the model's own quantities (`CENT`, `V`) and new hazard parameters
(`H0`, `beta`):

```{r tte-model-convert}
tm <- tte_model(my_model, hazard = "H0 * exp(lp + beta * (CENT/V))",
  params = list(H0 = 0.01, beta = 0.05))
tm
```

`tm$code` is the full converted model, ready to inspect:

```{r tte-model-code}
cat(tm$code, sep = "\n")
```

Pass `tm` to `sim_tte_ode()` exactly like a built-in model name:

```{r tte-model-sim}
sim_tm <- sim_tte_ode(model = tm, n = 30, end = 20, delta = 2, data = dose,
  seed = 1)
head(sim_tm$events)
```

If your own model already declares its own `$OMEGA` block (the classic
`TVCL`/`ETA()` idiom), `tte_model()` leaves it untouched and
`sim_tte_ode()`'s `omega` argument goes through it directly instead of
the `idata` route shown in section 3:

```{r tte-model-omega}
my_model_omega <- "
$PARAM TVCL = 1, TVV = 20, KA = 1
$OMEGA @labels ETA_CL ETA_V
0 0
$CMT DEPOT CENT
$MAIN
double CL = TVCL*exp(ETA_CL);
double V  = TVV*exp(ETA_V);
$ODE
dxdt_DEPOT = -KA*DEPOT;
dxdt_CENT = KA*DEPOT - (CL/V)*CENT;
"
tm_omega <- tte_model(my_model_omega,
  hazard = "H0 * exp(lp + beta * (CENT/V))",
  params = list(H0 = 0.01, beta = 0.05))
sim_tm_omega <- sim_tte_ode(model = tm_omega, omega = diag(c(0.09, 0.16)),
  n = 30, data = dose, end = 20, delta = 2, seed = 1)
head(sim_tm_omega$events)
```

A bare compiled `mrgsolve` model (not wrapped by `tte_model()`) is not
accepted directly by `sim_tte_ode()` -- convert it first. This keeps
there to exactly one way to bring a model in.

## 5. Accuracy, messages, and when things get slow

`delta` (and `add`) only control how finely the returned event time is
*reported* -- the event time itself is located inside the solver, at
its own internal step resolution, not snapped to the requested output
grid.

Two situations trigger a `message()` (never a change in what is
simulated, just a heads-up):

- **A very steep hazard at a coarse `delta`** can force the solver
  across a region where the hazard changes by orders of magnitude in
  one internal step; when that happens for some subjects, `sim_tte_ode()`
  falls back to a coarser-grid event-time estimate for just those
  subjects and says so, naming which of two known causes applies and
  which of `delta` or `rtol`/`atol` fixes it.
- **Weibull `shape < 0.05`**: the ODE form of the Weibull hazard is
  mathematically singular as $t \to 0$ for small shapes; below 0.05,
  accuracy degrades near $t = 0$ specifically. The simulation still
  runs; `sim_tte(type = "weibull")` (the closed-form path) is exact
  there if that matters more than the in-solver mechanism's other
  properties.

```{r shape-guardrail}
msgs <- capture.output(
  sim_small_shape <- sim_tte_ode(model = "weibull",
    param = list(mu = -1, shape = 0.01), n = 5, end = 10, delta = 1,
    seed = 1),
  type = "message"
)
msgs
```

`model = "mspline"` evaluates a quadratic M-spline basis *inside* the
compiled model as a genuinely continuous hazard, at the cost of a
fixed set of shipped interior-knot counts (3, 5, or 7); `sim_tte(type
= "ms")` accepts an arbitrary knot count but treats the hazard as
piecewise-constant on its own reporting grid instead. Use whichever
property you need more.

Every `tte_model()` conversion is a real C++ compile (a few seconds).
`mrgsolve::mcode_cache()` (used internally) skips recompiling an
unchanged, previously-seen model within the same cache directory --
`options(mrgsolve.project = <a persistent directory>)` makes that
cache survive across R sessions too, worthwhile if you reconvert the
same model many times (e.g. a simulation study script re-run
repeatedly). Two caveats: a caller-supplied `name` reused for
genuinely different `hazard`/`params` content, called back to back,
can in principle collide within the same build directory (leave
`name` at its default, which avoids this by construction); and in a
parallel workflow, give each worker its own cache directory, or simply
let each one recompile -- the compile cost is small next to the
parallel speed-up.

## 6. Backward compatibility

`sim_tte()` and `sim_tte_df()` -- the original, closed-form/grid-based
API -- are unchanged:

```{r legacy-api}
set.seed(42)
lp <- matrix(rnorm(20, 0, 0.5), nrow = 20)
result <- sim_tte(pi = lp, mu = -1, coefs = 1.1,
  time = seq(0.1, 20, by = 0.5), type = "weibull", end_time = 20)
head(result)
```

Reasons to still reach for them instead of `sim_tte_ode()`:

- **An arbitrary M-spline knot count** (`sim_tte(type = "ms")`) --
  `sim_tte_ode(model = "mspline")` ships only 3 fixed variants.
- **Exact Weibull behavior for very small shapes** (`shape < 0.05`,
  see section 5) -- the closed-form calculation has no singularity
  there.
- **An externally simulated `p11` trajectory** (from any source, not
  necessarily `mrgsolve`) -- `sim_tte_df()` accepts any validated
  `(ID, time, p11)` data frame directly; `sim_tte_ode()` always runs
  the ODE solver itself.
