---
title: "Simulating random (functional) censoring on top of an M-spline hazard"
subtitle: "Why a fixed study end time isn't the only kind of censoring worth simulating"
author: "Carlos Traynor"
date: "2026-07-14"
format: gfm
---

## Administrative vs. random censoring

In [the previous post](msplines-simulation.md) we simulated event times from
a flexible M-spline baseline hazard and let `sim_tte()` apply
**administrative censoring**: a single fixed cutoff (`end_time`) at which
the study simply stops, and anyone who hasn't had the event yet is
censored at that instant. This is the easiest kind of censoring to reason
about — every censored subject has exactly the same censoring time — and
it's a perfectly reasonable model for "the trial ended on this date."

But a lot of real censoring doesn't look like that. Patients drop out at
different times for different reasons — they move away, withdraw consent,
get lost to follow-up, or die of unrelated causes. None of that lines up
with a single administrative cutoff. The more general and more realistic
way to model this is **random (functional) censoring**: censoring time is
itself a random variable, drawn independently for each subject from its
own distribution, competing directly with the event time. Whichever
happens first — the event or the censoring — is what you actually observe.

This matters for simulation studies because administrative-only censoring
is a best case: every subject contributes information right up until the
same cutoff. Random censoring is messier and more realistic, and it's the
setting you need if you want to stress-test an estimator's behavior under
varying, unpredictable follow-up times rather than one clean cutoff.

## Step 1: true event times from the M-spline hazard

We'll reuse the exact M-spline specification from the previous post — same
knots, same degree-2 basis, same coefficients — so the two posts read as a
series. The one thing we change is the intercept `mu`: in the first post
we used `mu = -0.5`, which left a substantial fraction of subjects
administratively censored at the 24-unit horizon. Here we want genuine
*true* event times for (almost) everyone, uncontaminated by that
administrative boundary, so we can cleanly layer our own random censoring
on top afterward. Bumping the intercept up to `mu = 1.5` pushes essentially
every subject's event before the 24-unit horizon:

```{r}
#| label: setup
#| message: false
library(simtte)
library(splines2)
library(survival)

set.seed(20260714)
```

```{r}
#| label: true-event-times
time_new     <- seq(0.1, 24, by = 0.1)
knots_new    <- quantile(time_new, probs = c(0.15, 0.35, 0.55, 0.75))
boundary_new <- c(0, 24)
basis_new <- mSpline(time_new, knots = knots_new, Boundary.knots = boundary_new,
                      degree = 2, intercept = TRUE)
coefs_new <- c(0.05, 0.35, 0.55, 0.30, 0.10, 0.04, 0.02)
mu_event  <- 1.5   # vs. -0.5 in the M-splines post, so events resolve within the horizon

n <- 300
lp <- matrix(rep(0, n), nrow = n)
sim_true <- sim_tte(pi = lp, mu = mu_event, basis = basis_new, coefs = coefs_new,
                     time = time_new, type = "ms", end_time = 24)

event_time <- sim_true$sim_time
mean(sim_true$sim_status)  # fraction with a genuine event within the modeled horizon
```

With `mu_event = 1.5`, essentially everyone reaches their true event before
the spline's modeled horizon runs out — `event_time` is now a clean vector
of *true* event times, free of any administrative artifact.

## Step 2: an independent Weibull censoring time

Now we add a second, completely independent process: a censoring time
drawn per subject from a Weibull distribution, `rweibull(n, shape, scale)`.
The `shape` parameter controls whether dropout risk is roughly constant,
accelerating, or decelerating over time; `scale` controls the typical
timing of dropout. Holding `shape` fixed and varying `scale` is the
simplest way to dial the overall censoring rate up or down — a smaller
scale means earlier, more frequent censoring:

```{r}
#| label: tune-censoring-rate
shape_c <- 1.5
scale_candidates <- c("~50%" = 2.5, "~30%" = 4.7, "~10%" = 12.5)

for (lbl in names(scale_candidates)) {
  set.seed(1)
  cens <- rweibull(n, shape = shape_c, scale = scale_candidates[[lbl]])
  cat(lbl, "target -> scale =", scale_candidates[[lbl]],
      "-> achieved rate =", round(mean(cens < event_time), 3), "\n")
}
```

Three `scale` values, three very different censoring intensities, same
`shape`. That's the whole lever.

## Step 3: combining event and censoring times

This is the mechanical heart of random-censoring simulation, and it's
worth seeing explicitly rather than hidden in a helper function. For any
subject, we only ever *observe* the earlier of the two competing times,
plus a flag for which one it was:

```{r}
#| label: explicit-combine
cens_30 <- {
  set.seed(1)
  rweibull(n, shape = shape_c, scale = scale_candidates[["~30%"]])
}

observed_time <- pmin(event_time, cens_30)
status        <- as.integer(event_time <= cens_30)

head(data.frame(event_time, cens_30, observed_time, status))
mean(status == 0)  # achieved censoring rate
```

`status == 1` means we observed the true event; `status == 0` means
censoring happened first and the true event time (had we kept following
that subject) is unknown. Now that the mechanics are visible, we can wrap
it in a small reusable function and apply it across all three target
rates:

```{r}
#| label: apply-to-all-rates
apply_censoring <- function(event_time, shape, scale, seed = 1) {
  set.seed(seed)
  cens <- rweibull(length(event_time), shape = shape, scale = scale)
  data.frame(
    observed_time = pmin(event_time, cens),
    status        = as.integer(event_time <= cens)
  )
}

d50 <- apply_censoring(event_time, shape_c, scale_candidates[["~50%"]])
d30 <- apply_censoring(event_time, shape_c, scale_candidates[["~30%"]])
d10 <- apply_censoring(event_time, shape_c, scale_candidates[["~10%"]])

data.frame(
  target   = names(scale_candidates),
  scale    = unname(scale_candidates),
  achieved = c(mean(d50$status == 0), mean(d30$status == 0), mean(d10$status == 0))
)
```

## How censoring distorts what you can observe

The true event times are, by construction, fully observed (no censoring
at all), so a Kaplan-Meier fit on `event_time` directly gives us the true
underlying survival curve. Let's compare that to the Kaplan-Meier curve
you'd actually estimate from the *observed*, randomly censored data at
each of the three censoring rates:

```{r}
#| label: fig-comparison
#| fig-cap: "True (uncensored) survival curve vs. the Kaplan-Meier curve estimated from randomly censored data, at three censoring rates"
#| fig-width: 12
#| fig-height: 4.2
fit_true <- survfit(Surv(event_time, rep(1, n)) ~ 1)

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))

fit_50 <- survfit(Surv(d50$observed_time, d50$status) ~ 1)
plot(fit_true, col = "black", lwd = 2, conf.int = FALSE,
     xlab = "Time", ylab = "Survival probability", main = "~50% random censoring")
lines(fit_50, col = "#3366CC", lwd = 2, conf.int = FALSE)
legend("topright", legend = c("True", "Observed"), col = c("black", "#3366CC"),
       lwd = 2, bty = "n", cex = 0.8)

fit_30 <- survfit(Surv(d30$observed_time, d30$status) ~ 1)
plot(fit_true, col = "black", lwd = 2, conf.int = FALSE,
     xlab = "Time", ylab = "Survival probability", main = "~30% random censoring")
lines(fit_30, col = "#339966", lwd = 2, conf.int = FALSE)
legend("topright", legend = c("True", "Observed"), col = c("black", "#339966"),
       lwd = 2, bty = "n", cex = 0.8)

fit_10 <- survfit(Surv(d10$observed_time, d10$status) ~ 1)
plot(fit_true, col = "black", lwd = 2, conf.int = FALSE,
     xlab = "Time", ylab = "Survival probability", main = "~10% random censoring")
lines(fit_10, col = "#CC3333", lwd = 2, conf.int = FALSE)
legend("topright", legend = c("True", "Observed"), col = c("black", "#CC3333"),
       lwd = 2, bty = "n", cex = 0.8)
```

At ~10% censoring the observed curve tracks the truth closely across
almost the whole timeline. At ~30% it starts to diverge earlier and the
tail becomes visibly noisier. At ~50%, the observed Kaplan-Meier curve
departs from the truth much earlier and stops well short of the true
curve's range entirely — with half the sample censored, there simply
isn't enough information left to estimate the tail of the distribution,
and the step function gets choppier throughout. Same underlying M-spline
hazard, same sample size — the only thing that changed is how aggressively
an *independent* process competed with the event.

## Why this matters, and a caveat

Random censoring is the setting you want whenever a simulation study needs
to look like a real trial with imperfect follow-up rather than a clean,
uniformly-observed cohort. A few concrete uses:

- **Stress-testing an estimator's robustness** to different amounts and
  shapes of censoring — does a proposed survival model or test still
  perform well when 40-50% of subjects are censored, or does its variance
  blow up?
- **Mimicking realistic dropout patterns** in a trial design simulation —
  power calculations that assume administrative-only censoring can be
  optimistic if your actual population drops out steadily over time.
- **Comparing informative vs. non-informative censoring** — building
  simulations where censoring risk depends on covariates or even on the
  (unobserved) event risk itself, to see how badly standard methods break
  down when the non-informative-censoring assumption is violated.

That last point is also this post's caveat. Everything here used
**independent, non-informative censoring**: the Weibull censoring time was
drawn completely independently of the M-spline event time, for every
subject. Standard survival methods (Kaplan-Meier, Cox models, and `simtte`
itself) lean on that independence assumption. Real dropout is often
**informative** — sicker patients may be more likely to drop out, or more
likely to have the event, correlating the two processes — and when that's
true, naively treating censoring as independent will bias your estimates.
Simulating *that* kind of censoring means deliberately correlating the
censoring and event mechanisms (e.g., making the censoring hazard depend
on the same linear predictor as the event hazard) — a natural next step
once you're comfortable with the independent case shown here.

## References

- Bender R, Augustin T, Blettner M (2005). Generating survival times to
  simulate Cox proportional hazards models. *Statistics in Medicine*,
  24(11), 1713–1723. <https://doi.org/10.1002/sim.2059>
- Royston P, Parmar MKB (2002). Flexible parametric proportional-hazards
  and proportional-odds models for censored survival data, with
  application to prognostic modelling and estimation of treatment
  effects. *Statistics in Medicine*, 21(15), 2175–2197.
  <https://doi.org/10.1002/sim.1203>
