---
title: "Simulating flexible survival hazards with M-splines in `simtte`"
subtitle: "A hands-on walkthrough of building a custom M-spline baseline hazard for time-to-event simulation"
author: "Carlos Traynor"
date: "2026-07-14"
format: gfm
---

## Why M-splines, and why should you care?

If you simulate survival data for a living — clinical trial design, health
economics models, methods research — you've probably reached for the
Weibull distribution more times than you can count. It's a great default:
two parameters, a closed-form hazard, done. But real hazards are rarely
that tidy. Post-surgical risk that spikes early and fades. Cancer
recurrence risk that rises, plateaus, then declines. Multi-modal hazards
from mixed patient populations. A single Weibull shape parameter simply
can't reproduce these patterns.

This is where **M-splines** come in. An M-spline is a set of smooth,
*non-negative* basis functions built from piecewise polynomials joined at
"knots." Because each basis function is non-negative everywhere, any
non-negative linear combination of them is guaranteed to be non-negative
too — which is exactly the property you need to represent a baseline
*hazard* function (hazards can't go negative). Add enough basis functions
and place your knots thoughtfully, and you can approximate almost any
hazard shape you like, without committing to a rigid parametric form. This
is the same idea behind the flexible parametric survival models of
Royston & Parmar (2002).

`simtte` bakes this in: give it a spline basis matrix and a coefficient
vector, and it will treat their product as your baseline hazard, simulate
the implied cumulative hazard via `mrgsolve`, and draw event times by
inverse transform sampling (Bender et al., 2005). The package ships one
example M-spline specification (`ms_data`) so you can try the machinery
immediately — but it's just *one* hazard shape among infinitely many you
could build. Below, we build a different one from scratch.

## A quick look at the built-in example

`simtte`'s bundled `ms_data` dataset is a list with four pieces:

```{r}
#| label: setup
#| message: false
library(simtte)
library(splines2)
library(survival)

set.seed(20260714)
```

```{r}
#| label: builtin-example
data("ms_data")
str(ms_data)
```

`basis` is a 299 × 6 matrix — one row per time point, one column per
spline basis function — built with a **cubic (degree 3)** M-spline over a
`0`–`7.3` time horizon. Multiply `basis %*% coefs` and you get the
baseline hazard at each time point. That's the entire recipe.

## Building a new M-spline hazard from scratch

Let's design a hazard with a different flavor: instead of a short,
sharply-peaked risk window, imagine a 24-month oncology follow-up where
risk climbs gradually, plateaus for a while, and then tails off slowly —
a broader, gentler hump spread over a much longer horizon. We'll also use
a **quadratic** (degree 2) spline with **7** basis functions instead of 6,
just to show the specification is entirely up to you.

The basis matrix itself is built with `splines2::mSpline()` — the same
function used to construct the package's built-in `ms_data$basis` (you
can confirm this yourself: the `basis` attributes on `ms_data$basis`
record `degree`, `knots`, and `Boundary.knots`, and calling `mSpline()`
with those exact values reproduces it to floating-point precision).
`simtte` doesn't ship a basis-building helper itself — it expects you to
bring your own basis matrix, however you construct it, and just hands
back a hazard shaped by whatever coefficients you multiply it by.

```{r}
#| label: new-spec
## ---- New M-spline baseline hazard spec ----
time_new     <- seq(0.1, 24, by = 0.1)                     # 24-month horizon
knots_new    <- quantile(time_new, probs = c(0.15, 0.35, 0.55, 0.75))
boundary_new <- c(0, 24)

basis_new <- mSpline(
  time_new,
  knots          = knots_new,
  Boundary.knots = boundary_new,
  degree         = 2,             # quadratic, vs. cubic (degree 3) in ms_data
  intercept      = TRUE
)
coefs_new <- c(0.05, 0.35, 0.55, 0.30, 0.10, 0.04, 0.02)   # 7 basis functions, vs. 6
mu_new    <- -0.5

stopifnot(ncol(basis_new) == length(coefs_new))
dim(basis_new)
```

Four internal knots at the 15th/35th/55th/75th percentiles of the time
range, degree 2, and hand-picked coefficients that rise then decay — that's
the whole specification. Nothing here is copied from `ms_data`; it's a
genuinely different hazard shape, time grid, and basis dimension.

## Simulating from the new spec

With the basis matrix and coefficients in hand, simulating event times
looks exactly like the built-in example — `sim_tte()` doesn't care where
your basis came from:

```{r}
#| label: simulate-new
n <- 200
lp_new <- matrix(rep(0, n), nrow = n)   # no covariate effect, for a clean comparison

sim_new <- sim_tte(
  pi     = lp_new,
  mu     = mu_new,
  basis  = basis_new,
  coefs  = coefs_new,
  time   = time_new,
  type   = "ms"
)

head(sim_new)
mean(sim_new$sim_status)   # observed event rate
```

`sim_tte()` returns the familiar output: `sim_time`, `sim_status` (1 =
event, 0 = administratively censored at 24), `ID`, and `lp`.

## Comparing the new hazard to the built-in example

To see how different the two specifications really are, let's simulate
200 subjects under each (with a flat, zero linear predictor so we're
comparing baseline hazards apples-to-apples) and plot both the underlying
hazard shapes and the resulting Kaplan-Meier curves:

```{r}
#| label: fig-comparison
#| fig-cap: "Comparison of baseline hazard shapes and simulated Kaplan-Meier curves between the built-in ms_data example and the new M-spline specification"
#| fig-width: 9
#| fig-height: 4.2
data("ms_data")
lp_builtin <- matrix(rep(0, n), nrow = n)
sim_builtin <- sim_tte(
  pi = lp_builtin, mu = ms_data$mu, basis = ms_data$basis,
  coefs = ms_data$coefs, time = ms_data$time, type = "ms"
)

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

# (a) baseline hazard shapes
haz_builtin <- ms_data$basis %*% ms_data$coefs
haz_new     <- basis_new %*% coefs_new
plot(ms_data$time, haz_builtin, type = "l", lwd = 2, col = "#3366CC",
     xlab = "Time", ylab = "Baseline hazard", main = "Baseline hazard shapes",
     xlim = c(0, max(time_new)), ylim = range(c(haz_builtin, haz_new)))
lines(time_new, haz_new, lwd = 2, col = "#CC3333")
legend("topright", legend = c("Built-in ms_data", "New spec"),
       col = c("#3366CC", "#CC3333"), lwd = 2, bty = "n", cex = 0.8)

# (b) Kaplan-Meier comparison of simulated event times
fit_builtin <- survfit(Surv(sim_time, sim_status) ~ 1, data = sim_builtin)
fit_new     <- survfit(Surv(sim_time, sim_status) ~ 1, data = sim_new)
plot(fit_builtin, col = "#3366CC", lwd = 2, conf.int = FALSE,
     xlab = "Time", ylab = "Survival probability",
     main = "Simulated survival curves", xlim = c(0, max(time_new)))
lines(fit_new, col = "#CC3333", lwd = 2, conf.int = FALSE)
legend("topright", legend = c("Built-in ms_data", "New spec"),
       col = c("#3366CC", "#CC3333"), lwd = 2, bty = "n", cex = 0.8)
```

The difference is obvious at a glance. The built-in `ms_data` hazard
(blue) is squeezed into a short window — it spikes early, dips, and spikes
again, all within about 7 time units. Our new spec (red) spreads a single,
gentler hump across a 24-unit horizon, giving a Kaplan-Meier curve that
declines more gradually and levels off with more subjects still at risk of
an event later on. Same simulation engine, same `sim_tte()` call — a
completely different clinical story.

## Why this matters

It's easy to reach for a Weibull model because it's fast and familiar, but
if the hazard you're trying to simulate doesn't actually look like a
Weibull hazard, your "realistic" simulated trial isn't realistic at all —
it just *looks* clean. M-splines let you sketch the hazard shape you
actually believe in (early risk bump, long plateau, whatever your domain
knowledge tells you) and get exact, reproducible event-time simulations
out the other end, with the same one-line `sim_tte()` call you'd use for
Weibull data.

The practical takeaway: don't treat `ms_data` as *the* M-spline example —
treat it as a template. Swap in your own knots, degree, and coefficients,
and `simtte` will simulate whatever hazard shape your problem actually
calls for.

## References

- Bender R, Augustin T, Blettner M (2005). Generating survival times to
  simulate Cox proportional hazards models. *Statistics in Medicine*,
  24(11), 1713–1723. <https://doi.org/10.1002/sim.2059>
- Royston P, Parmar MKB (2002). Flexible parametric proportional-hazards
  and proportional-odds models for censored survival data, with
  application to prognostic modelling and estimation of treatment
  effects. *Statistics in Medicine*, 21(15), 2175–2197.
  <https://doi.org/10.1002/sim.1203>
