---
title: "Driving a hazard from a PK/PD ODE system in simtte"
subtitle: "Coupling a pharmacokinetic/pharmacodynamic model directly to the event hazard, instead of fitting a flexible parametric shape after the fact"
author: "Carlos Traynor"
date: "2026-07-14"
format: gfm
---

## When a flexible hazard still isn't the right hazard

The [first post](msplines-simulation.md) in this series showed how M-splines
let you sketch almost any hazard *shape* you like — early bumps, plateaus,
long tails — without committing to a parametric family. That's a huge step
up from a bare Weibull model, but it's still fundamentally **empirical**:
you choose knots and coefficients to produce a hazard shape that *looks*
plausible, then simulate from it.

Sometimes you know more than "the hazard looks roughly like this." If
you're simulating a trial for a drug with a known or hypothesized
mechanism of action — say, a compound that suppresses an inflammatory
mediator, and elevated levels of that mediator drive event risk — you
already have a mechanistic model for *why* the hazard should evolve the
way it does: it's downstream of the drug's pharmacokinetics (PK, how the
body handles the drug) and pharmacodynamics (PD, what the drug does to the
body). In that setting, the more faithful simulation isn't "pick spline
coefficients that resemble the expected effect" — it's coupling the
hazard directly to a PK/PD ODE system and letting the mechanism generate
the hazard shape itself, including all its cohort-level heterogeneity, as
a direct consequence of the pharmacology rather than a curve-fitting
exercise.

This post builds exactly that: a small PK/PD model in `mrgsolve`, coupled
to the survival probability as an ODE state, simulated the same way the
`simtte` package's own built-in Weibull and M-spline models are — and fed
straight into `sim_tte_df()`.

## The PK/PD model

We'll use a standard, well-established pharmacometric building block: a
one-compartment PK model with first-order oral absorption, feeding an
**indirect response model** (Dayneka et al. 1993) for the PD effect. The
drug doesn't act on the hazard instantaneously through concentration —
it suppresses the *production* of a mediator `R`, whose level accumulates
or decays according to its own turnover kinetics. The hazard then depends
on the current level of that mediator, not on concentration directly. This
"lag" between concentration and the risk-driving state is exactly the kind
of indirect, accumulating PD effect the introduction promised.

Concretely:

- PK: absorption rate constant $k_a$, clearance $CL$, volume $V$, giving
  concentration $C(t) = A_{\text{cent}}(t) / V$.
- PD: drug fractionally inhibits the production rate of mediator $R$,
  $$\mathrm{INH}(t) = \frac{I_{\max} \cdot C(t)}{IC_{50} + C(t)}, \qquad
  \frac{dR}{dt} = k_{in}\bigl(1 - \mathrm{INH}(t)\bigr) - k_{out} R,$$
  with $R(0) = k_{in}/k_{out}$ (the untreated steady state).
- Hazard: proportional to the *relative* level of the mediator versus its
  untreated baseline, $h(t) = h_0 \cdot R(t) / (k_{in}/k_{out})$, so
  $h(t) = h_0$ at baseline and falls as the drug suppresses $R$.

Rather than computing this hazard and integrating it after the fact, we
add the survival probability itself as a fourth ODE state, exactly the
way `simtte`'s own bundled Weibull and M-spline models do it internally
(`inst/models/ms.cpp`, `inst/models/weibull.cpp`): $dp_{11}/dt = -p_{11}
\cdot h(t)$, with $p_{11}(0) = 1$. `mrgsolve`'s ODE solver then produces
$p_{11}(t)$ directly, correctly integrated alongside the PK and PD states
in one pass -- no separate trapezoidal-integration step needed.

```{r}
#| label: setup
#| message: false
library(simtte)
library(mrgsolve)
library(dplyr)
library(survival)

set.seed(20260714)
```

```{r}
#| label: pkpd-model
code <- '
$PARAM CL = 2, V = 20, KA = 1, KIN = 1, KOUT = 0.5, IC50 = 1, IMAX = 0.9, H0 = 0.15

$CMT GUT CENT R

$INIT p11 = 1

$OMEGA @labels ECL EV EIC50
0.09 0.09 0.16

$MAIN
double CLi   = CL*exp(ECL);
double Vi    = V*exp(EV);
double IC50i = IC50*exp(EIC50);
R_0 = KIN/KOUT;

$ODE
double CP  = CENT/Vi;
double INH = IMAX*CP/(IC50i+CP);
double HAZ = H0*(R/(KIN/KOUT));
dxdt_GUT  = -KA*GUT;
dxdt_CENT = KA*GUT - (CLi/Vi)*CENT;
dxdt_R    = KIN*(1-INH) - KOUT*R;
dxdt_p11  = -p11*HAZ;

$CAPTURE CP HAZ
'
mod <- mcode("pkpd_hazard", code, quiet = TRUE)
mod
```

The model has between-subject variability (`$OMEGA`) on clearance, volume,
and PD potency (`IC50`) -- standard lognormal random effects, the usual
`mrgsolve` idiom -- so every simulated subject gets their own PK/PD (and
therefore hazard and survival) trajectory.

## Simulating and inspecting PK/PD trajectories first

Before generating a single event time, let's simulate a small cohort under
a once-daily oral dosing regimen (100 mg, 24 doses over a 24-day window)
and look at what's actually driving risk:

```{r}
#| label: fig-trajectories
#| fig-cap: "Individual pharmacodynamic hazard trajectories (left) and the corresponding survival probability p11 (right) for six simulated subjects under once-daily dosing"
#| fig-width: 9
#| fig-height: 4
n_preview <- 6
data_preview <- expand.ev(ID = 1:n_preview, amt = 100, cmt = 1, ii = 1, addl = 23, time = 0)
out_preview <- as.data.frame(mrgsim(mod, data = data_preview, end = 24, delta = 0.1))

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
plot(NULL, xlim = c(0, 24), ylim = range(out_preview$HAZ),
     xlab = "Time (days)", ylab = "Hazard", main = "PD-driven hazard")
for (i in 1:n_preview) {
  d <- out_preview[out_preview$ID == i, ]
  lines(d$time, d$HAZ, col = i)
}
plot(NULL, xlim = c(0, 24), ylim = c(0, 1),
     xlab = "Time (days)", ylab = "p11 (survival probability)", main = "Individual survival curves")
for (i in 1:n_preview) {
  d <- out_preview[out_preview$ID == i, ]
  lines(d$time, d$p11, col = i)
}
```

The hazard starts at its untreated baseline ($h_0 = 0.15$), falls sharply
over the first few days as the drug builds up and suppresses the
mediator, and levels off once concentration and mediator level reach a
new steady state under chronic dosing. The six subjects trace out
slightly different curves -- that's the between-subject variability on
clearance, volume, and potency doing its job.

## From the ODE output straight into `sim_tte_df()`

This is the part worth being precise about, since it's easy to get wrong
silently. `sim_tte_df()` expects a data frame with individual-level
(subject, time) rows and a survival-probability column named `p11` by
default (the probability of remaining event-free at that time -- *not* a
hazard). Because we built `p11` as a coupled ODE state above, the
`mrgsolve` output already contains it directly; there is no manual
hazard-to-survival conversion step required in R.

There is, however, an undocumented contract detail worth flagging: the
package's internal helpers pull the time value by **raw column
position 2** of the data frame, not by the column name `"time"`. We
confirmed this by testing it directly -- reordering columns so that
`p11` sits in position 2 silently returns *p11 values* mislabelled as
sampled event times, with no error. The only safe pattern is to hand
`sim_tte_df()` a data frame with columns in the exact order `(ID, time,
p11)`:

```{r}
#| label: sim-tte-df-demo
dat_for_sim <- out_preview %>% select(ID, time, p11)
head(dat_for_sim)

sim_preview <- sim_tte_df(dat_for_sim)
sim_preview
```

Six subjects in, six simulated event/censoring outcomes out, using exactly
the same `sim_tte_df()` call as the [custom mrgsolve output
example](../vignettes/advanced-usage.Rmd) in the package's own advanced
vignette -- the only difference is that our `p11` column came from a
PK/PD model instead of a hand-written toy survival probability.

## Simulating a full cohort with censoring

Now the same thing at trial scale: 300 subjects, and -- to build on [the
second post](functional-censoring-simulation.md) rather than starting
over -- an independent Weibull censoring time layered on top, using the
same mechanics (`observed_time = pmin(...)`, `status = 1(event <=
censoring)`), with the important twist that our event time here already
blends a *true* event with `sim_tte_df()`'s own administrative censoring
at the 24-day horizon, so the true-event flag has to be carried through
explicitly:

```{r}
#| label: full-cohort
n <- 300
data_full <- expand.ev(ID = 1:n, amt = 100, cmt = 1, ii = 1, addl = 23, time = 0)
out_full  <- as.data.frame(mrgsim(mod, data = data_full, end = 24, delta = 0.1))

sim_event <- sim_tte_df(out_full %>% select(ID, time, p11))
mean(sim_event$sim_status)  # event rate before any additional censoring

apply_censoring <- function(event_time, true_status, shape, scale, seed = 1) {
  set.seed(seed)
  cens <- rweibull(length(event_time), shape = shape, scale = scale)
  data.frame(
    observed_time = pmin(event_time, cens),
    status        = as.integer(true_status == 1 & event_time <= cens)
  )
}

cohort <- apply_censoring(sim_event$sim_time, sim_event$sim_status, shape = 1.5, scale = 20)
mean(cohort$status)  # event rate after adding independent random censoring
```

## Does the simulation recover the intended survival function?

The best sanity check for a mechanistic simulator is to compare the
Kaplan-Meier curve estimated from the simulated (censored) event times
against the population-average `p11` curve implied by the PK/PD model --
they should track each other closely, since under non-informative
censoring the Kaplan-Meier estimator is unbiased for the true marginal
survival function that `p11`, averaged over the between-subject
variability, is estimating:

```{r}
#| label: fig-km-vs-p11
#| fig-cap: "Kaplan-Meier curve from the simulated, censored cohort versus the population-average p11 curve implied by the PK/PD model"
#| fig-width: 7
#| fig-height: 4.5
fit_km <- survfit(Surv(cohort$observed_time, cohort$status) ~ 1)
mean_p11 <- out_full %>% group_by(time) %>% summarise(mean_p11 = mean(p11))

plot(fit_km, col = "#CC3333", lwd = 2, conf.int = FALSE,
     xlab = "Time (days)", ylab = "Survival probability",
     main = "Simulated KM vs. population-average p11")
lines(mean_p11$time, mean_p11$mean_p11, col = "black", lwd = 2, lty = 2)
legend("topright", legend = c("KM (observed, censored)", "Mean p11 (population-average)"),
       col = c("#CC3333", "black"), lwd = 2, lty = c(1, 2), bty = "n", cex = 0.8)
```

The two track closely, with a modest gap opening up in the tail -- exactly
what you'd expect from Monte Carlo noise at 300 subjects plus a shrinking
risk set at later times, rather than any systematic bias in the
simulation mechanics.

## Does the mechanistic link actually do anything?

To confirm the PK/PD coupling isn't decorative, we can re-run the
simulation at three different values of `IC50` (potency -- lower means a
more potent drug, since less concentration is needed to achieve half-maximal
inhibition) and check that the simulated event rate moves the way
pharmacology says it should:

```{r}
#| label: potency-sensitivity
for (ic50 in c(0.3, 1, 5)) {
  out_x <- as.data.frame(mrgsim(mod, data = data_full, end = 24, delta = 0.1,
                                 param = list(IC50 = ic50)))
  sim_x <- sim_tte_df(out_x %>% select(ID, time, p11))
  cat("IC50 =", ic50, "-> simulated event rate =", round(mean(sim_x$sim_status), 3), "\n")
}
```

A more potent drug (`IC50 = 0.3`) yields the lowest event rate, a weaker
one (`IC50 = 5`) the highest, with the reference value in between -- the
mechanistic link is doing real work, not just sitting there for show.

## Why this matters, and when it's worth the complexity

A coupled PK/PD hazard is considerably more work to set up than a Weibull
or M-spline model: you need a defensible PK/PD structure, plausible
parameter values (and their between-subject variability), and a dosing
regimen, on top of everything the earlier posts already covered. That
complexity buys you something a purely empirical hazard shape cannot:
every knob in the simulation maps onto something with a real
pharmacological interpretation, and it does so at the *individual* level.
That matters when:

- You're simulating a trial for a drug with a known or hypothesized
  mechanism, and want the simulated hazard shape to be a genuine
  consequence of that mechanism rather than a curve chosen to resemble it.
- You want to propagate PK/PD parameter uncertainty -- clearance
  variability, potency, onset -- through to trial-level outcomes such as
  power or expected event counts, rather than just varying an abstract
  hazard-shape parameter.
- You're exploring "what if" pharmacology questions (a more potent
  compound, faster onset, different dosing frequency) and want the
  resulting change in the survival curve to be an emergent consequence of
  the mechanism, as demonstrated in the potency comparison above.

If none of that mechanistic detail is actually informing your simulation
-- if you just need *a* plausible-looking hazard shape -- the M-spline
approach from the first post is far less work for a very similar result.
Save the PK/PD route for when the mechanism itself is the thing you're
trying to simulate.

## References

- Dayneka NL, Garg V, Jusko WJ (1993). Comparison of four basic models of
  indirect pharmacodynamic responses. *Journal of Pharmacokinetics and
  Biopharmaceutics*, 21(4), 457-478. <https://doi.org/10.1007/BF01061691>
- Bender R, Augustin T, Blettner M (2005). Generating survival times to
  simulate Cox proportional hazards models. *Statistics in Medicine*,
  24(11), 1713-1723. <https://doi.org/10.1002/sim.2059>
